---
title: "Rain gauge optimization v1"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Get the basin subcatchments for the upper Severn

```{r}
library(tidyverse)
library(httr2)
library(jsonlite)
library(tidyplots)
library(ggsci)
library(leaflet)
library(htmltools)
library(sf)
library(terra)
library(tidyterra)
library(ggspatial)
library(ncdf40)
library(RColorBrewer)
library(GGally)

```

## Load gis layers

```{r}
###--------------------------------------------------------

flow_stations_meta<-readRDS("flow_stations_meta.rda")

###load plynlimon

Plyn_catch <- read_sf("Headwater_catchments/Plyn_catchments/data/Shapefile/ee2eeee5-e456-4e93-87e9-eee97ee149ee/PlynlimonCatchments.shp", crs=st_crs(27700))

###Load river_net Plyn

Plyn_RN <- read_sf("Headwater_catchments/Plyn_rivernet/data/Shapefile/PlynlimonRiverNetwork.shp" ,crs=st_crs(27700))


###Load raster with 2023 LC data from CEH (10 m res)----------------------------------

catchment_LMC<-terra::rast(x = "lmc2023 clip.tif")
#All_LMC<-terra::rast(x = "C:/Users/khamisk/Dropbox/LCM_2023_10m/lcm-2023-10m_5816930/gblcm2023_10m.tif" )

### Load DEM -----------------------------------------

dem_Dolwen_1m <- terra::rast("dem_Dolwen_1m.tif")
dem_Dolwen <- terra::rast("dem_Dolwen.tif")
dem_wider <- terra::rast("dem_Dolwen_wider.tif")

#### load other shapefiles
catchment_Dolwen <- read_sf("54080/54080.shp")
river_net_wider <- read_sf("54080_river_net/StreamLine_54080.shp"
                           ,crs=st_crs(27700))
river_net_Dolwen = read_sf("river_net_Dolwen.shp")
river_net_CEH <- read_sf("River_net/DigitalRiverNetwork_50K_GB.shp",
                         crs=st_crs(27700))

#### Load soil info--------------------------------

BGS_soil_shp<-read_sf("BGS_soil_parent.shp")

BGS_HOST<-rast("HOST.tif")

BGS_HOST<-terra::project(BGS_HOST,"EPSG:27700")

BGS_soil_Dolwen<-st_intersection(x = BGS_soil_shp,y = catchment_Dolwen)

BGS_HOST_Dolwen <- terra::crop(BGS_HOST,catchment_Dolwen,mask=T)

#### clip plyn

Plyn_RN_S = st_intersection(Plyn_RN, catchment_Dolwen)

Plyn_catch_S<-Plyn_catch |>  filter( Catchment=="Severn")

Plyn_RN_S_10k = st_intersection(river_net_Dolwen, Plyn_catch_S )


```

## Load the locations for gauge and weather station

```{r}

plyn_sites<-read_csv("Headwater_catchments/Plyn_sites_meta.csv")


plyn_sites_bng <- st_as_sf(plyn_sites, coords = c("x","y"),crs=27700)


```

## rainfall analysis

```{r}

####
rain_meta_cmd<-read_csv("rain_meta_for_CMD.csv")

rain_meta_sf<-st_as_sf(x = rain_meta_cmd, coords = c("x_2", "y_2"), crs=27700) ## load location of rain gauges in the catchment

###load rainfall

ukgrid      <- "+init=epsg:27700"
ukgrid2      <- "epsg:27700"
latlong     <- "+init=epsg:4326"

proj        <-ukgrid 
HAD_names<-dir(path = "HAD_rainfall_annual/",full.names = T,pattern = ".nc")###check file path works!

HAD_ann<-terra::rast(HAD_names)
crs(HAD_ann)<-ukgrid2

###crop to catchment

HAD_Dolwen_ppn <- terra::crop(HAD_ann,catchment_Dolwen,mask=T)
#HAD_Dolwen_ppn <- terra::crop(HAD_ann,grid_1km)

HAD_Dolwen_ppn_mean<-mean(HAD_Dolwen_ppn)
#HAD_Dolwen_ppn_range<-max(HAD_Dolwen_ppn) - min(HAD_Dolwen_ppn)

##plot to check
plot(HAD_Dolwen_ppn_mean)

###create a shapefile polygon grid
grid_1km <- HAD_Dolwen_ppn_mean %>% 
  st_make_grid(cellsize = 1000, what ="polygons") %>% # grid of points
  st_intersection(catchment_Dolwen)  


grid_1km_point <- 
HAD_Dolwen_ppn_mean %>% 
  st_make_grid(cellsize = 1000, what ="centers") %>% # grid of points
  st_intersection(catchment_Dolwen)  



  
###get points
grid_1km_xy<-st_coordinates(grid_1km_point)
#st_coordinates(grid_1km_sf)

##convert to simple features
grid_1km_sf<-st_sf(geom = grid_1km )

##plot grid to check
plot(grid_1km_xy)
 
 
```

## Calculate terrain variables

```{r}
###calc terrain metrics
slope_ras<-terrain(dem_Dolwen, v="slope", neighbors=8) 
aspect_ras<-terrain(dem_Dolwen, v="aspect", neighbors=8) 
tri_ras<-terrain(dem_Dolwen, v="TRI", neighbors=8) ##Terrain Ruggedness Index see help fro more info



dem_max<-terra::extract(dem_Dolwen[[1]],grid_1km_sf,fun=max)
dem_mean<-terra::extract(dem_Dolwen[[1]],grid_1km_sf,fun=mean)
dem_min<-terra::extract(dem_Dolwen[[1]],grid_1km_sf,fun=min)
dem_var<-terra::extract(dem_Dolwen[[1]],grid_1km_sf,fun=var)
slope_mean<-terra::extract(slope_ras,grid_1km_sf,fun=mean)
slope_var<-terra::extract(slope_ras,grid_1km_sf,fun=var)
aspect_mean<-terra::extract(aspect_ras,grid_1km_sf,fun=mean)
aspect_var<-terra::extract(aspect_ras,grid_1km_sf,fun=var)
tri_mean<-terra::extract(tri_ras,grid_1km_sf,fun=mean)
tri_var<-terra::extract(tri_ras,grid_1km_sf,fun=var)
PPN_mean<-terra::extract(HAD_Dolwen_ppn_mean,grid_1km_sf,fun=mean)

###plot variables against x and y


grid_1km_sf$max_elev<-dem_max$spat_3e42c6a1e8b_996_o8up7yuDdJ6XHBX
grid_1km_sf$min_elev<-dem_min$spat_3e42c6a1e8b_996_o8up7yuDdJ6XHBX
grid_1km_sf$var_elev<-dem_var$spat_3e42c6a1e8b_996_o8up7yuDdJ6XHBX
grid_1km_sf$mean_elev<-dem_mean$spat_3e42c6a1e8b_996_o8up7yuDdJ6XHBX
grid_1km_sf$range_elev<-dem_max$spat_3e42c6a1e8b_996_o8up7yuDdJ6XHBX - dem_min$spat_3e42c6a1e8b_996_o8up7yuDdJ6XHBX
grid_1km_sf$mean_slope<-slope_mean$slope
grid_1km_sf$var_slope<-slope_var$slope
grid_1km_sf$mean_aspect<-aspect_mean$aspect
grid_1km_sf$var_aspect<-aspect_var$aspect
grid_1km_sf$mean_tri<-tri_mean$TRI
grid_1km_sf$var_tri<-tri_var$TRI
grid_1km_sf$mean_PPN<-PPN_mean$mean


```

##plot to assess relationship between variables

```{r}
### extract cells that match the point grid

point_1km_sf<-st_intersection(grid_1km_sf,grid_1km_point)  


point_1km_sf$x<-grid_1km_xy[,1]
point_1km_sf$y<-grid_1km_xy[,2]


rain_meta_sf2<- st_intersection(grid_1km_sf,rain_meta_sf)

rain_meta_sf2<-rain_meta_sf2[1:12]

rain_xy<-st_coordinates(rain_meta_sf2)

rain_meta_sf2$x<-rain_xy[,1]
rain_meta_sf2$y<-rain_xy[,2]
 
###Check it matches

 ggplot()+geom_sf(data=rain_meta_sf, colour = "red",size=3)+  scale_fill_viridis_c(option = "magma")+geom_sf(data=river_net_Dolwen)+tidyterra::geom_spatraster(data = HAD_Dolwen_ppn_mean)+geom_sf(data= grid_1km,fill=NA)



```

## Scatter plot matrix for CMD

```{r}
##scatter plot matrix for relationships 
p <- drop_na(point_1km_sf) |> ggpairs(columns = c(4,6,8,12,13,14))+ggpubr::theme_pubr()+theme(axis.text.x = element_text(angle = 90)) 
 
p 
```

###Box plots of env. range and rain gauge location

```{r}

all_df_long<-drop_na(point_1km_sf[c(4,6,8,12,13,14,15)]) |>  st_drop_geometry() |> pivot_longer(cols = 1:6, names_to = "var", values_to = "value")

site_df_long<-rain_meta_sf2[c(4,6,8,12,13,14,15)] |>  st_drop_geometry() |> pivot_longer(cols = 1:6, names_to = "var", values_to = "value")

all_df_long |> ggplot() + geom_jitter(aes(x = var ,y = value),width=0.2,alpha=0.6, colour="gray")+facet_wrap(~var,scales="free") + geom_jitter(data = site_df_long,aes(x = var ,y = value),width=0.2,colour="red",size=1.7)+facet_wrap(~var,scales="free")+ggpubr::theme_pubr()+xlab("")+ylab("")+
  theme(strip.text.x = element_blank())
```

### Set up grid

This is messy and needs a function to be written...

```{r}

####create 6 x 6 design (NOTE can increase decrease the resolution here)

###rescale all the variables

point_1km_sf$mean_PPN_s<-scales::rescale(point_1km_sf$mean_PPN)
point_1km_sf$x_s<-scales::rescale(point_1km_sf$x)
point_1km_sf$y_s<-scales::rescale(point_1km_sf$y)
point_1km_sf$elev_s<-scales::rescale(point_1km_sf$mean_elev)
point_1km_sf$slope_s<-scales::rescale(point_1km_sf$mean_slope)

## create the points
a<-seq(min(point_1km_sf$x_s), max(point_1km_sf$x_s), length.out = 6)
a2<-seq(min(point_1km_sf$y_s), max(point_1km_sf$y_s), length.out =6) 
b<-seq(min(point_1km_sf$mean_PPN_s), max(point_1km_sf$mean_PPN_s), length.out = 6)
c<-seq(min(point_1km_sf$elev_s), max(point_1km_sf$elev_s), length.out = 6)
#d<-seq(min(point_1km_sf$slope_s), max(point_1km_sf$slope_s), length.out = 6)

##expand them to a grid
grid_x_ppn_s <- expand.grid(x = a, mean_PPN = b)
grid_y_ppn_s <- expand.grid(y = a2, mean_PPN = b)
grid_elev_ppn_s <- expand.grid(elev_s = c, mean_PPN_s = b)

 
###Do the same for raw data

a<-seq(min(point_1km_sf$x), max(point_1km_sf$x), length.out = 6)
a2<-seq(min(point_1km_sf$y), max(point_1km_sf$y), length.out =6) 
b<-seq(min(point_1km_sf$mean_PPN), max(point_1km_sf$mean_PPN), length.out = 6)
c<-seq(min(point_1km_sf$mean_elev), max(point_1km_sf$mean_elev), length.out = 6)
grid_x_ppn <- expand.grid(x = a, mean_PPN = b)
grid_y_ppn <- expand.grid(y = a2, mean_PPN = b)
grid_elev_ppn <- expand.grid(elev = c, mean_PPN = b)
 

```

##PPN and x cord

```{r}
#######--------------------------------x coord and PPN mm---------------------------- 


### create df  
poten <-tibble(x_s=point_1km_sf$x_s,mean_PPN_s =point_1km_sf$mean_PPN_s,site=rep("poten",times=length(point_1km_sf$x_s))) ## potential sites
 
ideal<-tibble(x_s=grid_x_ppn_s$x,mean_PPN_s = grid_x_ppn_s$mean_PPN,site=rep("ideal",times=length(grid_x_ppn_s$x))) ##ideal sites
 

df_combo<-rbind(poten,ideal)

###calc distance matrix--------------------------

distances <- dist(df_combo[c("x_s", "mean_PPN_s")], diag = TRUE, upper = TRUE)
distances <- as.matrix(distances)

distances <- distances[df_combo$site == "ideal",]
distances <-as.matrix(distances) 

distances <- distances[,1:nrow(poten)]

####loop to find min--------------------------

location<-rep(NA,times=nrow(ideal))

for(i in 1:nrow(ideal)){ 
location[i]<-which.min(distances[i,])
}
 
min_value<-rep(NA,times=nrow(ideal))
for(i in 1:nrow(ideal)){ 
min_value[i]<-min(distances[i,])
}

###set cutoff for distance and filter to remove sites above threshold

df_selected<-tibble(location,min_value) |> filter(min_value<0.12)

df_combo$selected<-rep(1:length(df_combo$x_s))
point_1km_sf$selected<-rep(1:length(point_1km_sf$x_s))

df_combo<-df_combo |> mutate(selected_x_ppn=case_when( selected %in% df_selected$location ~ "T",
                                        .default = "NA"))

point_1km_sf<-point_1km_sf |> mutate(selected_x_ppn=case_when( selected %in% df_selected$location ~ "T",
                                        .default = "NA"))

df_chosen<-df_combo |> filter(selected_x_ppn=="T")
 

df_chosen_sf  <- point_1km_sf |>  filter(selected_x_ppn=="T") 


##Plot in standardised units

p1<-ggplot()+geom_point(data=point_1km_sf,aes(x_s,mean_PPN_s),colour="gray",size=2)+
  geom_point(data=grid_x_ppn_s,aes(x,mean_PPN),color="red",shape=3)+ggpubr::theme_pubr()+ 
  geom_point(data = df_chosen , aes(x_s,mean_PPN_s), shape = 21, size = 3) +
  xlab("X Coordinate") +ylab("Annual P")
  


##in original units

p2<-ggplot()+geom_point(data=point_1km_sf,aes(x,mean_PPN),colour="gray",size=2)+
  geom_point(data=grid_x_ppn,aes(x,mean_PPN),color="red", shape=3)+ggpubr::theme_pubr()+ 
  geom_point(data=rain_meta_sf2,aes(x,mean_PPN),colour="black",size=4)+
  geom_point(data =df_chosen_sf  , aes(x,mean_PPN), shape = 21, size = 3) +
  xlab("X Coordinate") +ylab("Annual P (mm)")


```

## y and PPN mm

```{r}


#######--------------------------------y and PPN mm---------------------------- 
### create df  
poten <-tibble(y_s=point_1km_sf$y_s,mean_PPN_s =point_1km_sf$mean_PPN_s,site=rep("poten",times=length(point_1km_sf$y_s)))
 
ideal<-tibble(y_s=grid_y_ppn_s$y,mean_PPN_s = grid_y_ppn_s$mean_PPN,site=rep("ideal",times=length(grid_y_ppn_s$y)))
 

df_combo<-rbind(poten,ideal)

###calc distance matrix

distances <- dist(df_combo[c("y_s", "mean_PPN_s")], diag = TRUE, upper = TRUE)
distances <- as.matrix(distances)

distances <- distances[df_combo$site == "ideal",]
distances <-as.matrix(distances) 

distances <- distances[,1:nrow(poten)]

####loop to find min

location<-rep(NA,times=nrow(ideal))

for(i in 1:nrow(ideal)){ 
location[i]<-which.min(distances[i,])
}
 
min_value<-rep(NA,times=nrow(ideal))
for(i in 1:nrow(ideal)){ 
min_value[i]<-min(distances[i,])
}

df_selected<-tibble(location,min_value) |> filter(min_value<0.12)

df_combo$selected<-rep(1:length(df_combo$y_s))
point_1km_sf$selected<-rep(1:length(point_1km_sf$y_s))

df_combo<-df_combo |> mutate(selected_y_ppn=case_when( selected %in% df_selected$location ~ "T",
                                        .default = "NA"))

point_1km_sf<-point_1km_sf |> mutate(selected_y_ppn=case_when( selected %in% df_selected$location ~ "T",
                                        .default = "NA"))

df_chosen_y<-df_combo |> filter(selected_y_ppn=="T")
 

df_chosen_sf_y  <- point_1km_sf |>  filter(selected_y_ppn=="T") 


##Plot in standardised units

p3<-ggplot()+geom_point(data=point_1km_sf,aes(y_s,mean_PPN_s),colour="gray",size=2)+
  geom_point(data=grid_y_ppn_s,aes(y,mean_PPN),color="red",shape=3)+ggpubr::theme_pubr()+ 
  geom_point(data = df_chosen_y , aes(y_s,mean_PPN_s), shape = 21, size = 3) +
  xlab("Y Coordinate") +ylab("Annual P")
  


##in original units

p4<-ggplot()+geom_point(data=point_1km_sf,aes(y,mean_PPN),colour="gray",size=2)+
  geom_point(data=grid_y_ppn,aes(y,mean_PPN),color="red", shape=3)+ggpubr::theme_pubr()+ 
  geom_point(data=rain_meta_sf2,aes(y,mean_PPN),colour="black",size=4)+
  geom_point(data =df_chosen_sf_y  , aes(y,mean_PPN), shape = 21, size = 3) +
  xlab("Y Coordinate") +ylab("Annual P (mm)")



```

## elev and PPN mm

```{r}
#######--------------------------------elev and PPN mm---------------------------- 
### create df  
poten <-tibble(elev_s=point_1km_sf$elev_s,mean_PPN_s =point_1km_sf$mean_PPN_s,site=rep("poten",times=length(point_1km_sf$y_s)))
 
ideal<-tibble(elev_s=grid_elev_ppn_s$elev_s,mean_PPN_s = grid_elev_ppn_s$mean_PPN,site=rep("ideal",times=length(grid_y_ppn_s$y)))
 

df_combo<-rbind(poten,ideal)

###calc distance matrix

distances <- dist(df_combo[c("elev_s", "mean_PPN_s")], diag = TRUE, upper = TRUE)
distances <- as.matrix(distances)

distances <- distances[df_combo$site == "ideal",]
distances <-as.matrix(distances) 

distances <- distances[,1:nrow(poten)]

####loop to find min

location<-rep(NA,times=nrow(ideal))

for(i in 1:nrow(ideal)){ 
location[i]<-which.min(distances[i,])
}
 
min_value<-rep(NA,times=nrow(ideal))
for(i in 1:nrow(ideal)){ 
min_value[i]<-min(distances[i,])
}

df_selected<-tibble(location,min_value) |> filter(min_value<0.12)

df_combo$selected<-rep(1:length(df_combo$elev_s))
point_1km_sf$selected<-rep(1:length(point_1km_sf$elev_s))

df_combo<-df_combo |> mutate(selected_elev_ppn=case_when( selected %in% df_selected$location ~ "T",
                                        .default = "NA"))

point_1km_sf<-point_1km_sf |> mutate(selected_elve_ppn=case_when( selected %in% df_selected$location ~ "T",
                                        .default = "NA"))

df_chosen_elev<-df_combo |> filter(selected_elev_ppn=="T")
 

df_chosen_sf_elev  <- point_1km_sf |>  filter(selected_elve_ppn=="T") 


##Plot in standardised units

p5<-ggplot()+geom_point(data=point_1km_sf,aes(elev_s,mean_PPN_s),colour="gray",size=2)+
  geom_point(data=grid_elev_ppn_s,aes(elev_s,mean_PPN_s),color="red",shape=3)+ggpubr::theme_pubr()+ 
  geom_point(data = df_chosen_elev , aes(elev_s,mean_PPN_s), shape = 21, size = 3) +
  xlab("Elevation") +ylab("Annual P")
  


##in original units

p6<-ggplot()+geom_point(data=point_1km_sf,aes(mean_elev,mean_PPN),colour="gray",size=2)+
  geom_point(data=grid_elev_ppn,aes(elev,mean_PPN),color="red", shape=3)+ggpubr::theme_pubr()+ 
  geom_point(data=rain_meta_sf2,aes(mean_elev,mean_PPN),colour="black",size=4)+
  geom_point(data =df_chosen_sf_elev  , aes(mean_elev,mean_PPN), shape = 21, size = 3) +
  xlab("Elevation (m)") +ylab("Annual P (mm)")









```

## Combined plots

```{r}
cowplot::plot_grid(p2,p4,p6)
```

## Maps of site optimization

```{r}


 ggplot()+geom_sf(data= grid_1km_sf,aes(fill = mean_PPN))+geom_sf(data=rain_meta_sf, colour = "red",size=3)+ scale_fill_grass_c(palette = "blues") +theme_light() +xlab("")+ylab("") +
    annotation_scale(
    height = unit(0.015, "npc"),
    width_hint = 0.5,
    pad_x = unit(0.07, "npc"),
    pad_y = unit(0.03, "npc"),
    text_cex = .8
  ) +
  annotation_north_arrow(
    which_north = TRUE,
    pad_x = unit(0.84, "npc"),
    pad_y = unit(0.83, "npc"),
    style = north_arrow_fancy_orienteering()
  ) +theme(plot.margin = unit(c(0,0,0,0), "cm")) +labs(fill="Ann. P (mm)") + geom_sf(data=df_chosen_sf_elev,size=3) + geom_sf(data=df_chosen_sf_y, size = 3, shape = 3)+ geom_sf(data=df_chosen_sf, size = 3, shape = 7)

```
